CrypTool Script\footnote{ Background reading for the CrypTool program}\\ Mathematics and Cryptography

### 2003-06-27 script/en/chapters/cryptomethods.tex
be updates

Up-to-the-minute procedure is the symmetric AES standard. The associated
Rijndael algorithm was declared winner of the AES award on 2 October 2000
and thus succeedes the DES procedure.

More details about the AES algorithms can be found within the 
online help of CrypTool\index{CrypTool}%
\footnote{%
      CrypTool online help\index{CrypTool}: the index head-word {\bf AES}
      leads to the 3 help pages: {\bf AES candidates}, 
      {\bf The AES winner Rijndael} and 
      {\bf The Rijndael encryption algorithm}.
  }.

@@

AES with a minimum key length of 128 bit is still in the long run sufficiently secure against brute-force attacks -- as long as the quantum computers aren't powerful enough. When announced as new standard AES was immune against all known crypto attacks, mostly based on statistical considerations and earlier applied to DES: using pairs of clear and cipher texts expressions are constructed, which are not completely at random, so they allow conclusions to the used keys. These attacks required unrealistically large amounts of intercepted data.

@@

The mathematicians Courtois and Pieprzyk \cite{Courtois2002} published a paper
in 2002, which got a great deal of attention amongst the crypto community: The
pair had further developed the XL-method (eXtended Linearization), introduced at
Eurocrypt 2000 by Shamir et al., to create the so called XSL-method (eXtended
Sparse Linearization). The XL-method is a heuristic technique, which in some
cases manages to solve big non-linear equation systems and which was till then
used to analyze an asymmetric algorithm (HFE).  The innovation of Courtois and
Pieprzyk was, to apply the XL-method on symmetric codes: the XSL-method can be
applied to very specific equation systems. A 256-bit AES could be attacked in
roughly $2^{230}$ steps. This is still a purely academic attack, but also a
direction pointer for a complete class of block ciphers. The major problem with
this attack is that until now nobody has worked out, under what conditions it is
successful: the authors specify in their paper necessary conditions, but it is
not known, which conditions are sufficient.  There are two very new aspects of
this attack: firstly this attack is not based on statistics but on algebra. So
attacks seem to be possible, where only very small amounts of cipher text are
available. Secondly the security of a product-algorithm does not exponentially
increase with the number of rounds.

@@

So for 256-AES the attack is much more effective than brute-force but still far more away from any computing power which could be accessible in the short-to-long term.

@@

% --------------------------------------------------------------------------
\subsubsection{Current status of brute-force attacks on 
symmetric algorithms (RC5)}
\index{Brute-force attack}
\index{RC5}
\label{Brute-force-gegen-Symmetr}

@@

Similar cipher challenges are there for asymmetric algorithms (please see chapter 3.11.4).

@@

The most well-known asymmetric procedure is the \index{RSA} 
RSA algorithm\index{CrypTool}%
\footnote{%
The RSA algorithm is extensively described in chapter 3.10 and later
within this script. 
The RSA cryptosystem can be executed in many variations with 
CrypTool\index{CrypTool} v1.3 (using the menu 
{\bf Individual Procedures \textbackslash{} RSA Demonstration \textbackslash{}
RSA Cryptosystem}). 
The topical research results concerning RSA are described in chapter 3.11.
}%
,
named after its developers Ronald \index{Rivest Ronald} Rivest, Adi
\index{Shamir Adi} Shamir and Leonard \index{Adleman Leonard} Adleman. The RSA algorithm
was published in 1978. The concept of asymmetric encryption was first
introduced by Whitfield Diffie \index{Diffie Whitfield}  and Martin
\index{Hellman Martin} Hellman in 1976. Today, the ElGamal \index{ElGamal}
procedures also play a decisive role, particularly the \index{Schnorr C.P.} Schnorr
variant in the \index{DSA} DSA (Digital \index{Signature!digital}Signature
Algorithm).

>>>>

Up-to-the-minute procedure is the symmetric AES standard. The associated
Rijndael algorithm was declared winner of the AES award on October 2nd, 2000
and thus succeedes the DES procedure.

More details about the AES algorithms and the AES candidates of the last round
can be found within the online help of CrypTool\index{CrypTool}%
\footnote{%
      CrypTool online help\index{CrypTool}: the index head-word {\bf AES}
      leads to the 3 help pages: {\bf AES candidates}, 
      {\bf The AES winner Rijndael} and 
      {\bf The Rijndael encryption algorithm}.
  }.

@@

AES with a minimum key length of 128 bit is still in the long run sufficiently secure against brute-force attacks\index{Attack!brute-force} -- as long as the quantum computers aren't powerful enough. When announced as new standard AES was immune against all known crypto attacks, mostly based on statistical considerations and earlier applied to DES: using pairs of clear and cipher texts expressions are constructed, which are not completely at random, so they allow conclusions to the used keys. These attacks required unrealistically large amounts of intercepted data.

@@

The mathematicians Courtois and Pieprzyk \cite{Courtois2002} published a paper
in 2002, which got a great deal of attention amongst the crypto community: The
pair had further developed the XL-method (eXtended Linearization), introduced at
Eurocrypt 2000 by Shamir et al., to create the so called XSL-method (eXtended
Sparse Linearization). The XL-method is a heuristic technique, which in some
cases manages to solve big non-linear equation systems and which was till then
used to analyze an asymmetric algorithm (HFE).  The innovation of Courtois and
Pieprzyk was, to apply the XL-method on symmetric codes: the XSL-method can be
applied to very specific equation systems. A 256-bit AES could be attacked in
roughly $2^{230}$ steps. This is still a purely academic attack, but also a
direction pointer for a complete class of block ciphers. The major problem with
this attack is that until now nobody has worked out, under what conditions it is
successful: the authors specify in their paper necessary conditions, but it is
not known, which conditions are sufficient.  There are two very new aspects of
this attack: firstly this attack is not based on statistics but on algebra. So
attacks seem to be possible, where only very small amounts of cipher text are
available. Secondly the security of a product-algorithm\index{Cascade cipher}%
\footnote{%
A cipher text can be used as input for another encryption algorithm. 
A cascade cipher is build up as a composition of different encryption 
transformations.\\
Cascading does not always improve the security.\\
This process is also used within modern algorithms:
They usually combine simple and, considered at its own, cryptologically
relatively unsecure single steps in several rounds into an efficient 
overall procedure.  Most block ciphers (e.g. DES, IDEA) are cascade ciphers.\\
Also serial usage of the same cipher with different keys (like with Triple-DES)
is called cascade cipher. 
}
does not exponentially increase with the number of rounds.

@@

So for 256-AES the attack is much more effective than brute-force\index{Attack!brute-force} but still far more away from any computing power which could be accessible in the short-to-long term.

@@

% --------------------------------------------------------------------------
\subsubsection{Current status of brute-force attacks on 
symmetric algorithms (RC5)}
\index{Attack!brute-force}
\index{RC5}
\label{Brute-force-gegen-Symmetr}

@@

Similar cipher challenges are there for asymmetric algorithms (please see chapter \ref{NoteFactorisation}).

@@

The most well-known asymmetric procedure is the \index{RSA} 
RSA algorithm\index{CrypTool}%
\footnote{%
The RSA algorithm is extensively described in chapter \ref{rsabeweis} and later
within this script. 
The RSA cryptosystem can be executed in many variations with 
CrypTool\index{CrypTool} v1.3 (using the menu 
{\bf Individual Procedures \textbackslash{} RSA Demonstration \textbackslash{}
RSA Cryptosystem}). 
The topical research results concerning RSA are described 
in chapter \ref{SecurityRSA}.
}%
,
named after its developers Ronald \index{Rivest Ronald} Rivest, Adi
\index{Shamir Adi} Shamir and Leonard \index{Adleman Leonard} Adleman. The RSA algorithm
was published in 1978. The concept of asymmetric encryption was first
introduced by Whitfield Diffie \index{Diffie Whitfield}  and Martin
\index{Hellman Martin} Hellman in 1976. Today, the ElGamal \index{ElGamal}
procedures also play a decisive role, particularly the \index{Schnorr C.P.} Schnorr
variant in the \index{DSA} DSA (Digital \index{Signature!digital}Signature
Algorithm).

### 2003-06-27 script/en/chapters/ellipticcurves.tex
be updates

In many business sectors secure and efficient data transfer is essential.
In particular, the RSA algorithm is used in many applications. Although the
security of RSA is beyond doubt, the evolution in computing power has
caused a growth in the necessary key length.  Today, 1024-bit RSA keys are
standard, but the GISA\index{GISA} (German Information Security Agency)
recommends the usage of 2048-bit keys from 2006 on (compare
section~\ref{SecurityRSA}).  The fact that most chips on smart cards cannot
process keys extending 1024 bit shows that there is a need for
alternatives. Elliptic curve cryptography (ECC) can be such an
alternative in the field of asymmetric cryptography.

@@

Anyway that's enough about the role of elliptic curves in pure mathematics. 
In 1985 Neal Koblitz\index{Koblitz Neal} and Victor Miller\index{Miller Victor}
independently suggested using elliptic curves in cryptography. Elliptic 
curves have thus also found a concrete practical application. 
Another interesting field of application for elliptic curves is for
factorising whole numbers \index{Factorisation} (the RSA cryptographic system
is based on the \index{Complexity} difficulty/complexity of finding prime
factors of an extremely large number;  compare section \ref{SecurityRSA}.).
In this area, procedures based on elliptic curves have been investigated and
used since 1987 (compare section \ref{ECC-Factorisation}). \\
There are also prime number tests\index{Prime number!test} based on elliptic
curves.
% (s. Anmerkung U. Kuehn)

@@

Fields may contain an infinite number of elements (e.g. the field of real numbers). They are called {\em infinite} fields. In contrast we call a field
{\em finite}, if it contains only a finite number of elements (e.g. ${\mathbb Z}_p = \{0, 1, 2, 3, \cdots, p-1\}$
, where $p$ is a prime. ${\mathbb Z}_p$ with addition mod $p$ and multiplication
mod $p$).
\index{Field!Characteristic}
\paragraph{Characteristic of a field}
Let $K$ be a field and $1$ be the neutral element of $K$ with
respect to the multiplicative operation ``$\cdot$''. Then the characteristic of $K$ is said to be the order of $1$ with respect to the additive operation. This means that the characteristic of $K$ is the smallest positive integer $n$ such that
$$ \underbrace{1+1+\dots+1}_{\hbox{$n$ times}} =0 .
$$
If there is no such $n$, i.e. if $1+1+\dots+1\ne 0$ no matter how many $1$s we add, then we call $K$ a field
with characteristic $0$.

@@

\index{Finite!Fields}
\paragraph{Finite Fields}
As mentioned above, each finite field has a characteristic $p\ne 0$, where $p$ is a prime. On the other hand, given a prime $p$ there is a field which has exactly $p$ elements, that is ${\mathbb Z}_p$.

@@

Given a secure curve, the time that is needed to solve the ECDLP is strongly correlated with parameter $p$ in case $GF(p)$ respectively $n$ in case of $GF(2^n)$. The larger these parameters become, the more time an attacker needs to solve the ECDLP --- at least with the best algorithms known so far. Experts recommend bit-lengths of $200$ for $p$ for secure curves. A comparison with RSA modulus length shows why elliptic curves are so interesting for applications. We note that the computation effort for signing and encryption is closely related to the bit-length of the parameters. In addition the initiation process, i.e. the generation of the private-public-key-pair, becomes more complicated the larger $p$ is. Thus, one looks for the smallest parameters that still come along with the security required. It is remarkable that a length of $200$ bits for $p$ is sufficient to construct a {\em good} elliptic curve that is as secure as RSA with a $1024$ bit \index{RSA!Modulus} RSA-Modulus (as far as we know today). For short, the reason for this advantage of ECC lies in the fact that the best algorithms known for solving the ECDLP need exponential time while the best algorithms for factorizing are sub-exponential (number field sieve, quadratic sieve or factorizing with elliptic curves). Hence, the parameters for a cryptosystem that is based on the problem of {\em factorizing large integers} have to be larger than the parameters for a system based on ECDLP.

@@

\begin{sloppypar}
  The {\em elliptic curve discrete logarithm
    problem}\index{Problem of discrete logarithm} (ECDLP) \index{ECDLP} is the basis for elliptic curve cryptography. Based on this problem, there are different signature schemes. In order to apply one of these, we need:
\end{sloppypar}
\begin{itemize}
    \item An elliptic curve {\bf E} with an underlying field $GF(p^n)$.
    \item A prime $q\ne p$ and a point $G$ on the elliptic curve ${\bf E}$ with order $q$. This means that $qG={\cal O}$ and $rG\ne {\cal O}$ for all $r\in \{1,2,\dots,q-1\}$. Thus $q$ is a factor of the group order (i.e. the number of elements) $\#{\bf E}$ of $E$. Since $q$ is prime, $G$ generates a cyclic sub-group of ${\bf E}$ of order $q$.
\end{itemize}
The parameters mentioned are often called \index{Domain-Parameter} {\em Domain}-Parameter. They describe the elliptic curve ${\bf E}$ and the cyclic sub-group of ${\bf E}$ on which the signature scheme is based.

>>>>

In many business sectors secure and efficient data transfer is essential.
In particular, the RSA algorithm is used in many applications. Although the
security of RSA is beyond doubt, the evolution in computing power has
caused a growth in the necessary key length.  Today, 1024-bit RSA keys are
standard, but the GISA\index{GISA} (German Information Security Agency)
recommends the usage of 2048-bit keys from 2006 on (compare
section~\ref{SecurityRSA}).  The fact that most chips on smart cards cannot
process keys extending 1024 bit shows that there is a need for
alternatives. Elliptic curve cryptography (ECC) can be such an
alternative in the area of asymmetric cryptography.

@@

Anyway that's enough about the role of elliptic curves in pure mathematics. 
In 1985 Neal Koblitz\index{Koblitz Neal} and Victor Miller\index{Miller Victor}
independently suggested using elliptic curves in cryptography. Elliptic 
curves have thus also found a concrete practical application. 
Another interesting area of application for elliptic curves is for
factorising whole numbers \index{Factorisation} (the RSA cryptographic system
is based on the \index{Complexity} difficulty/complexity of finding prime
factors of an extremely large number;  compare section \ref{SecurityRSA}.).
In this area, procedures based on elliptic curves have been investigated and
used since 1987 (compare section \ref{ECC-Factorisation}). \\
There are also prime number tests\index{Prime number!test} based on elliptic
curves.
% (s. Anmerkung U. Kuehn)

@@

Fields may contain an infinite number of elements (e.g. the field of real numbers). They are called {\em infinite} fields. In contrast we call a field
{\em finite}, if it contains only a finite number of elements (e.g. ${\mathbb Z}_p = \{0, 1, 2, 3, \cdots, p-1\}$
, where $p$ is a prime. ${\mathbb Z}_p$ with addition mod $p$ and multiplication
mod $p$).
\index{Field!characteristic}
\paragraph{Characteristic of a field}
Let $K$ be a field and $1$ be the neutral element of $K$ with
respect to the multiplicative operation ``$\cdot$''. Then the characteristic of $K$ is said to be the order of $1$ with respect to the additive operation. This means that the characteristic of $K$ is the smallest positive integer $n$ such that
$$ \underbrace{1+1+\dots+1}_{\hbox{$n$ times}} =0 .
$$
If there is no such $n$, i.e. if $1+1+\dots+1\ne 0$ no matter how many $1$s we add, then we call $K$ a field
with characteristic $0$.

@@

\index{Field!finite}
\paragraph{Finite Fields}
As mentioned above, each finite field has a characteristic $p\ne 0$, where $p$ is a prime. On the other hand, given a prime $p$ there is a field which has exactly $p$ elements, that is ${\mathbb Z}_p$.

@@

Given a secure curve, the time that is needed to solve the ECDLP is strongly correlated with parameter $p$ in case $GF(p)$ respectively $n$ in case of $GF(2^n)$. The larger these parameters become, the more time an attacker needs to solve the ECDLP --- at least with the best algorithms known so far. Experts recommend bit-lengths of $200$ for $p$ for secure curves. A comparison with RSA modulus length shows why elliptic curves are so interesting for applications. We note that the computation effort for signing and encryption is closely related to the bit-length of the parameters. In addition the initiation process, i.e. the generation of the private-public-key-pair, becomes more complicated the larger $p$ is. Thus, one looks for the smallest parameters that still come along with the security required. It is remarkable that a length of $200$ bits for $p$ is sufficient to construct a {\em good} elliptic curve that is as secure as RSA with a $1024$ bit \index{RSA!modulus} RSA modulus (as far as we know today). For short, the reason for this advantage of ECC lies in the fact that the best algorithms known for solving the ECDLP need exponential time while the best algorithms for factorizing are sub-exponential (number field sieve, quadratic sieve or factorizing with elliptic curves). Hence, the parameters for a cryptosystem that is based on the problem of {\em factorizing large integers} have to be larger than the parameters for a system based on ECDLP.

@@

\begin{sloppypar}
  The {\em elliptic curve discrete logarithm
    problem}\index{Problem of discrete logarithm} (ECDLP) \index{ECDLP} is the basis for elliptic curve cryptography. Based on this problem, there are different signature schemes. In order to apply one of these, we need:
\end{sloppypar}
\begin{itemize}
    \item An elliptic curve {\bf E} with an underlying field $GF(p^n)$.
    \item A prime $q\ne p$ and a point $G$ on the elliptic curve ${\bf E}$ with order $q$. This means that $qG={\cal O}$ and $rG\ne {\cal O}$ for all $r\in \{1,2,\dots,q-1\}$. Thus $q$ is a factor of the group order (i.e. the number of elements) $\#{\bf E}$ of $E$. Since $q$ is prime, $G$ generates a cyclic sub-group of ${\bf E}$ of order $q$.
\end{itemize}
The parameters mentioned are often called \index{Domain parameter}
{\em Domain} parameter. They describe the elliptic curve ${\bf E}$ and 
the cyclic sub-group of ${\bf E}$ on which the signature scheme is based.

### 2003-06-27 script/en/chapters/introduction.tex
be updates

This edition of the script mainly updates the summaries of the following 
topical research areas:
\vspace{-7pt}
\begin{itemize}
  \item the search for the largest prime numbers (generalized Mersenne 
        and Fermat primes) \\ 
%        and Fermat primes, ``M-40'') 
	(chap. \ref{spezialzahlentypen}, \ref{zahlentyp_mersenne}), 
  \item progress in number theory (``Primes in P'') (chap. \ref{PrimesinP}), 
  \item the factorisation of big numbers (TWIRL, RSA-160)\index{TWIRL device} 
        (chap. \ref{TWIRLDevice}, \ref{RSA-160}),
  \item progress in cryptanalysis of the AES standard 
        (chap. \ref{NeueAES-Analyse}) and
  \item progress in brute-force attacks against symmetric algorithms 
        (chap. \ref{Brute-force-gegen-Symmetr}).
\end{itemize}

@@

Because the articles in this script are largely self-contained, this text
can also be read independently of CrypTool\index{CrypTool}.

>>>>

This edition of the script mainly updates the summaries of the following 
topical research areas:
\vspace{-7pt}
\begin{itemize}
  \item the search for the largest prime numbers (generalized Mersenne 
        and Fermat primes) \\ 
%        and Fermat primes, ``M-40'') 
	(chap. \ref{spezialzahlentypen}, \ref{zahlentyp_mersenne}), 
  \item progress in number theory (``Primes in P'') (chap. \ref{PrimesinP}), 
  \item the factorisation of big numbers (TWIRL, RSA-160)\index{TWIRL device} 
        (chap. \ref{TWIRLDevice}, \ref{RSA-160}),
  \item progress in cryptanalysis of the AES standard 
        (chap. \ref{NeueAES-Analyse}) and
  \item progress in brute-force attacks against symmetric algorithms 
        (chap. \ref{Brute-force-gegen-Symmetr}).\index{Attack!brute-force}
\end{itemize}

@@

The articles in this script are largely self-contained and
can also be read independently of CrypTool\index{CrypTool}.

### 2003-06-27 script/en/chapters/numbertheory.tex
be updates

\newpage
%\hypertarget{Kap_3}
\section{Introduction to Elementary Number Theory with Examples}
(Bernhard Esslinger, July 2001, Updates: Dec. 2001, June 2002, May 2003) \\

@@

For more details on prime numbers (e.g. how ``Fermat's Little Theorem'' can be used to test extremely large 
numbers to determine whether they are prime), please refer to the article on \hyperlink{Kapitel_2}{prime numbers} in this script.

@@

% --------------------------------------------------------------------------
\subsubsection{Patterns and structures}
\hypertarget{kap_3_5_2}
In general mathematicians investigate ``Structures''\index{Structure}. They ask e.g. at $ a * x \equiv b $ mod $m$,
which values $x$ can take for given values of $a, ~b, ~m.$

@@

Cryptographically not only the unique nature of the inverse is important, but
also that the set of possible values has been exhausted.
\begin{theorem}\label{thm-zth-exhperm}
For $a,i\in \{1, \cdots, m-1 \}$ with ${\rm~gcd} (a,m)=1),$ then the product $a*i {\rm ~mod ~} m $ takes for a 
certain number $a$ all values from $\{1, \cdots ,m-1 \} $ (exhaustive permutation \index{Permutation} of the length $m-1$)\footnote{See
also theorem \ref{thm-zth-ordp} in \hyperlink{Kap_3_9}{chapter 3.9 Multiplicative order and primitive roots}.}.
\end{theorem}
\vskip +10 pt
The following three examples\footnote{%
See \hyperlink{AppArith1}{Appendix D of this chapter} for the source code to compute
the tables using Mathematica\index{Mathematica} and Pari-GP.\index{Pari-GP}
}
illustrate the properties of multiplicative inverses.

@@

% --------------------------------------------------------------------------
\vskip +10pt
\subsubsection{Fast calculation of high powers} \hypertarget{hohpot}{} \label{hohpot}\index{Power}
%\subsubsection*{3.6.3.1 ~~~Fast calculation of high powers} \hypertarget{hohpot}{} \index{Power}
%\addcontentsline{toc}{subsubsection}{~~~~~~~~~~3.6.3.1 ~Fast calculation of high powers}
RSA encryption and decryption\footnote{See chapter \ref{rsaproof} (Proof of the
RSA procedure with Euler-Fermat) \index{RSA} and chapter~\ref{rsaconcrete} (The
RSA procedure with actual numbers).} entails calculating high powers modulo
$m$.  For example, the calculation ($100^5) {\rm ~(mod~} 3)$ exceeds the 32-bit
long integer\index{Long integer} number range provided we calculate $a^n$ by
actually multiplying a with itself $n$ times in line with the definition. In
the case of extremely large numbers, even a fast computer chip would take
longer than the age of the universe to calculate a single exponential. Luckily,
there is an extremely effective shortcut for calculating exponentials (but not
for calculating logarithms).

@@

The unique inverse is an essential condition for cryptography (see section~\ref{rsaproof}).

@@

% --------------------------------------------------------------------------
\subsubsection{Patterns and structures}
\hypertarget{Kap_3_8}{}\label{patternsandstructures}
As mathematicians investigate the structure \index{Structure} $a *x \equiv b$ mod $m$ (see \hyperlink{kap_3_5_2}{chapter 3.5.2}), so they are interested in
the structure $ x^{a} \equiv b$ mod $m.$

@@

% --------------------------------------------------------------------------
\subsubsection{The theorem of Euler-Fermat}\index{Euler Leonhard}\index{RSA}
In order to prove the RSA procedure, we need Fermat's theorem and its 
generalisation (Euler-Fermat theorem) -- 
\hyperlink{KleinerSatzFermat-chap2}{please also see chapter 2.5}.

@@

The problems of \index{Factorisation} factorising\index{Factorisation!factorisation problem} $n=pq$ with $q\neq p$ and finding $J(n)$ have a similar degree of difficulty 
and if we find a solution for one of the two problems, we also have a solution for the
other\footnote{If we know the factors of $n=p*q$ with $p\neq q$, then $J(n)=(p-1)*(q-1) = n-(p+q)+1$. Additionally the factors $p$ and $q$ are
solutions of the quadratic equation
$ x^2 - (p+q) x + pq=0.
$\\
If only $n$ and $J(n)$ are known, then it is:
$pq=n$ and $p+q= n-J(n)+1.$ So you get $p$ and $q$ by solving the equation
$$ x^2 + (J(n)-n-1)x +n= 0.
$$} (please compare requisition 3 in section 3.10.1).

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\subsection{Proof of the RSA procedure with Euler-Fermat}
\hypertarget{Kapitel_3_10}{}\index{RSA}\label{rsaproof}

@@

% --------------------------------------------------------------------------
\subsubsection{Basic idea of public key cryptography}

>>>>

\newpage
%\hypertarget{Kap_3}
\hypertarget{Kapitel_3}{}

\section{Introduction to Elementary Number Theory with Examples}
(Bernhard Esslinger, July 2001, Updates: Dec. 2001, June 2002, May 2003) \\
\label{Kapitel_3}

@@

For more details on prime numbers (e.g. how ``Fermat's Little Theorem'' can
be used to test extremely large numbers to determine whether they are prime),
please refer to the article on \hyperlink{Kapitel_2}{\ref{Label_Kapitel_2}, 
Prime numbers} in this script.

@@

% --------------------------------------------------------------------------
\subsubsection{Patterns and structures}
\hypertarget{kap_3_5_2}{}
\label{Label_Kapitel_3_5_2}

In general mathematicians investigate ``Structures''\index{Structure}. They ask e.g. at $ a * x \equiv b $ mod $m$,
which values $x$ can take for given values of $a, ~b, ~m.$

@@

Cryptographically not only the unique nature of the inverse is important, but
also that the set of possible values has been exhausted.
\begin{theorem}\label{thm-zth-exhperm}
For $a,i\in \{1, \cdots, m-1 \}$ with ${\rm~gcd} (a,m)=1),$ then the product $a*i {\rm ~mod ~} m $ takes for a 
certain number $a$ all values from $\{1, \cdots ,m-1 \} $ (exhaustive permutation \index{Permutation} of the length $m-1$)\footnote{See
also theorem \ref{thm-zth-ordp} in \hyperlink{Kap_3_9}
{chapter \ref{MultOrdPrimitveRoot}, Multiplicative order and primitive roots}.}.
\end{theorem}
\vskip +10 pt
The following three examples\footnote{%
See \hyperlink{AppArith1}{Appendix D of this chapter} for the source code to compute
the tables using Mathematica\index{Mathematica} and Pari-GP.\index{Pari-GP}
}
illustrate the properties of multiplicative inverses.

@@

% --------------------------------------------------------------------------
\vskip +10pt
\subsubsection{Fast calculation of high powers} \hypertarget{hohpot}{} \label{hohpot}\index{Power}
%\subsubsection*{3.6.3.1 ~~~Fast calculation of high powers} \hypertarget{hohpot}{} \index{Power}
%\addcontentsline{toc}{subsubsection}{~~~~~~~~~~3.6.3.1 ~Fast calculation of high powers}
RSA encryption and decryption\footnote{See chapter \ref{rsabeweis} (Proof of the
RSA procedure with Euler-Fermat) \index{RSA} and chapter~\ref{rsaconcrete} (The
RSA procedure with actual numbers).} entails calculating high powers modulo
$m$.  For example, the calculation ($100^5) {\rm ~(mod~} 3)$ exceeds the 32-bit
long integer\index{Long integer} number range provided we calculate $a^n$ by
actually multiplying a with itself $n$ times in line with the definition. In
the case of extremely large numbers, even a fast computer chip would take
longer than the age of the universe to calculate a single exponential. Luckily,
there is an extremely effective shortcut for calculating exponentials (but not
for calculating logarithms).

@@

The unique invertibility\index{Invertibility} is an essential condition for
cryptography (see section~\ref{rsabeweis}).

@@

% --------------------------------------------------------------------------
\subsubsection{Patterns and structures}
\hypertarget{Kap_3_8}{}\label{patternsandstructures}
As mathematicians investigate the structure \index{Structure}
$a *x \equiv b$ mod $m$ (see \hyperlink{kap_3_5_2}
{chapter \ref{Label_Kapitel_3_5_2}}), so they are interested in
the structure $ x^{a} \equiv b$ mod $m.$

@@

% --------------------------------------------------------------------------
\subsubsection{The theorem of Euler-Fermat}\index{Euler Leonhard}\index{RSA}
\label{Label_KleinerSatzFermat-chap3}
In order to prove the RSA procedure, we need Fermat's theorem and its 
generalisation (Euler-Fermat theorem) -- 
\hyperlink{KleinerSatzFermat-chap2}{please see chapter \ref{primality_tests}}.

@@

The problems of \index{Factorisation} factorising\index{Factorisation!factorisation problem} $n=pq$ with $q\neq p$ and finding $J(n)$ have a similar degree of difficulty 
and if we find a solution for one of the two problems, we also have a solution for the
other\footnote{If we know the factors of $n=p*q$ with $p\neq q$, then $J(n)=(p-1)*(q-1) = n-(p+q)+1$. Additionally the factors $p$ and $q$ are
solutions of the quadratic equation
$ x^2 - (p+q) x + pq=0.
$\\
If only $n$ and $J(n)$ are known, then it is:
$pq=n$ and $p+q= n-J(n)+1.$ So you get $p$ and $q$ by solving the equation
$$ x^2 + (J(n)-n-1)x +n= 0.
$$} (please compare requisition 3 in \hyperlink{Kapitel_3_10_1}
{section \ref{Label_Kapitel_3_10_1}}).

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\subsection{Proof of the RSA procedure with Euler-Fermat}
\hypertarget{Kapitel_3_10}{}\index{RSA}\label{rsabeweis}

@@

% --------------------------------------------------------------------------
\subsubsection{Basic idea of public key cryptography}
\index{Cryptography!public key}
\hypertarget{Kapitel_3_10_1}{}
\label{Label_Kapitel_3_10_1}
